Retrofitting Effect Handlers onto OCaml

### 2020-11-01 retro-concurrency.tex
Updates

@@

\subsection{Exception handlers}

@@

In order to forward exceptions across C frames, the C stub function
|caml_start_program|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception the innermost OCaml exception handler (|raised_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
very common to use them for \emph{local} control flow.

\subsection{DWARF unwind tables}

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.

\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
% gdb ./meander.exe
(gdb) break caml_raise_exn
Breakpoint 1 at 0x3e5dc
(gdb) r
Starting program: meander.exe
Breakpoint 1, 0x925dc in caml_raise_exn ()
(gdb) bt
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_start_program ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:14
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_start_program ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}

\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.

@@

\subsection{Static semantics}

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
    | (E s) -> (fun k1 -> e1)
    | (F f) -> (fun k2 -> e2)
    | e -> (fun k -> match perform e with
            | v -> continue k v
            | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
the outer handler and returns the value and exceptions back.

@@

\if{0}
\begin{figure}
  \begin{mathpar}
    \begin{array}{rc}
      \textsc{StepC} & \inferrule
                          {(\term, \env, \fl, \ostack) \cstep \config}
                          {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} \\ \\
      \textsc{StepO} & \inferrule
                          {(\term, \env, k, \cstack) \ostep \config}
                          {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
  \caption{Top-level Reductions -- $\config \step \config$.}
  \label{sem:top}
\end{figure}
\fi

\begin{figure}
  \begin{mathpar}
    \begin{array}{cc}
      \textsc{StepC} ~
      \inferrule{(\term, \env, \fl, \ostack) \cstep \config}
                {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
      \textsc{StepO} ~
      \inferrule{(\term, \env, k, \cstack) \ostep \config}
                {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
  \caption{Top-level Reductions -- $\config \step \config$.}
  \label{sem:top}
\end{figure}

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions (Figure~\ref{sem:top}) can be
performed by either by taking a C step $\cstep$ or an OCaml step $\ostep$.
Intuitively, this splits the world into executing C code and OCaml code.

@@

OCaml reductions are given in Figure~\ref{sem:ostep}. In OCaml, reductions
always occur on the top-most fiber in the current stack. \textsc{AdminO}
performs administrative reductions. \textsc{CallO} evaluates an OCaml function
on the current stack. \textsc{ExtCall} captures the behaviour of external
calls. On an external call, we create a C stack with no frames, whose tail is
the current OCaml stack, and evaluate the C functions body in the new stack.
\textsc{RetToC} returns a value to the enclosing C stack. We have exactly one
fiber on the stack, and this was created in the rule \textsc{Callback}, whose
handler has identity return case alone and the environment is empty.
\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
body of the return case.

The rule \textsc{Handle} installs a handler by pushing a fiber with no frames
and the given handler. The rule \textsc{ExHn} handles an exception, if the
current handler has a matching exception case. Observe that fiber on which the
exception was raise has been unwound. The rule \textsc{ExFwdC} forwards the
exception to C. Here, there is exactly one fiber on the current stack, and the
handler does not have a matching exception case. The rule \textsc{ExFwdFib}
forwards the exception to the next fiber if the current handler does not handle
it.

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation.
\textsc{EffUnHn} models unhandled effect. If the effect bubbles up to the top
fiber, which we know does not have an effect case (see \textsc{Callback} rule),
we raise \textsc{unhandled} exception at the point where the corresponding
effect was performed. This is achieved by appending the captured continuation
to the front of the current continuation.

Observe that |continue| and |discontinue| are not part of the expressions. They
are encoded as follows:
\begin{smathpar}
  \begin{array}{rcl}
    \textsf{continue} ~k ~e & = & k ~(\olam{x}{x}) ~e \\
    \textsf{discontinue} ~k ~l ~e & = & k ~(\olam{x}{\throw{l}{x}}) ~e
  \end{array}
\end{smathpar}

Intuitively, resuming a continuation in both the cases involves evaluating the
appropriate abstraction on top of the continuation. We perform the
administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
given continuation to the front of the current continuation, and evaluates the
body of the closure.

@@

>>>>

\subsection{External calls and callbacks}
\label{sec:external}

@@

\subsection{Exception handlers}
\label{sec:exn_handlers}

@@

% Note |caml_call_ocaml| is |caml_start_program| in ocaml-multicore source

In order to forward exceptions across C frames, the C stub function
|caml_call_ocaml|, which is the entry point for both the main OCaml code and
callbacks into OCaml, pushes an exception handler frame that forwards the
exception the innermost OCaml exception handler (|raised_exn_c| in
Figure~\ref{fig:stock_stack}), or prints a fatal error (|fatal_uncaught|) if
there are no enclosing handlers. Exceptions are so cheap in OCaml that it is
very common to use them for \emph{local} control flow.

\subsection{Stack unwinding}

OCaml generates \emph{stack maps} in order to accurately identify roots on the
stack for assisting GC. For every call point in the program, OCaml compiler
emits the size of the frame and the set of all live registers in the frame
containing pointers into the heap. During a GC, the OCaml stack is walked and
the roots are marked.

OCaml also generates precise DWARF unwind information for OCaml, thanks to
which debuggers such as |gdb| and |lldb|, and profilers such as |perf| work
out-of-the-box. For example, for the program in Figures~\ref{code:meander_ml}
and~\ref{code:meander_c}, one could set a break point in |gdb| at
|caml_raise_exn| to get the following backtrace.
\begin{lstlisting}[language=c,basicstyle=\ttfamily\footnotesize]
% gdb ./meander.exe
(gdb) break caml_raise_exn
Breakpoint 1 at 0x3e5dc
(gdb) r
Starting program: meander.exe
Breakpoint 1, 0x925dc in caml_raise_exn ()
(gdb) bt
#0  0x925dc in caml_raise_exn ()
#1  0x6fd3e in camlMeander__c_to_ocaml_83 () at meander.ml:5
#2  0x925a4 in caml_call_ocaml ()
#3  0x8a84a in caml_callback_exn (...) at callback.c:145
#4  caml_callback (...) at callback.c:199
#5  0x76e0a in ocaml_to_c (unit=1) at meander_.c:5
#6  0x6fd77 in camlMeander__omain_88 () at meander.ml:10
#7  0x6fe92 in camlMeander__entry () at meander.ml:14
#8  0x6f719 in caml_program ()
#9  0x925a4 in caml_call_ocaml ()
#10 0x92e4c in caml_startup_common (...) at startup_nat.c:162
#11 0x92eab in caml_startup_exn (...) at startup_nat.c:167
#12 caml_startup (...) at startup_nat.c:172
#13 0x6f55c in main (...) at main.c:44
\end{lstlisting}
\noindent which corresponds to the stack layout in
Figure~\ref{fig:stock_stack}.

The same backtrace can also be obtained by using \emph{frame pointers} instead
of DWARF unwind tables. OCaml allows compiling code with frame pointers, but
they are not enabled by default. OCaml stack tends to be deep with small frames
due to the pervasive use of recursive functions, not all of which are
tail-recursive. Hence, the addition of frame pointers can significantly
increase the size of the
stack~\footnote{https://github.com/ocaml/ocaml/issues/5721\#issuecomment-472965549}.
Moreover, not using frame pointers saves two instructions in the function
prologue and epilogue, and makes an extra register (|rbp| on x86) available.
Note that the DWARF unwind information is complementary to the information used
by OCaml to walk the stack for GC.

@@

\subsection{Static semantics}
\label{sec:static_semantics}

@@

For handling the effects, our implementation extends OCaml's |match ... with|
syntax with effect patterns. The expression
\begin{lstlisting}
match e with
| None -> false | Some b -> b
| effect (E s) k1 -> e1 | effect (F f) k2 -> e2
\end{lstlisting}
\noindent is translated to the equivalent of
\begin{lstlisting}
match_with (fun () -> e)
{ retc = (function None -> false | Some b -> b);
  effc = (function
  | (E s) -> (fun k1 -> e1)
  | (F f) -> (fun k2 -> e2)
  | e -> (fun k -> match perform e with
          | v -> continue k v
          | exception e -> discontinue k e)); }
\end{lstlisting}
For the sake of exposition, we introduce a |('a,'b) handler| type. This handler
handles a |'a comp| that returns a |'a| value, and itself returns a |'b| value.
The handler has a return case of type |'a -> 'b|. The effect case |effc|
handles effects of type |'c eff| with |('c,'b) continuation| and returns a
value of type |'b|. The last pattern in |effc| forwards any unmatched effect to
the outer handler and returns the value and exceptions back to the original
performer.

@@

The initial configuration for an expression $e$ is
$\configa{e}{\emptyset}{\cstacka{[]}{\bullet}}$, where the environment and the
stack are empty. The top-level reductions
  \begin{mathpar}
    \begin{array}{cc}
      \textsc{StepC} ~
      \inferrule{(\term, \env, \fl, \ostack) \cstep \config}
                {\configa{\term}{\env}{\cstacka{\fl}{\ostack}} \step \config} &
      \textsc{StepO} ~
      \inferrule{(\term, \env, k, \cstack) \ostep \config}
                {\configa{\term}{\env}{\ostacka{k}{\cstack}} \step \config}
    \end{array}
  \end{mathpar}
can be performed by either by taking a C step $\cstep$ or an OCaml step
$\ostep$. Intuitively, this splits the world into executing C code and OCaml
code.

@@

OCaml reductions are given in Figure~\ref{sem:ostep}. In OCaml, reductions
always occur on the top-most fiber in the current stack. \textsc{AdminO}
performs administrative reductions. \textsc{CallO} evaluates an OCaml function
on the current stack. \textsc{ExtCall} captures the behaviour of external
calls. On an external call, we create a C stack with no frames, whose tail is
the current OCaml stack, and evaluate the C functions body in the new stack.
\textsc{RetToC} returns a value to the enclosing C stack. In this case, we have exactly one
fiber on the stack, and this was created in the rule \textsc{Callback}, whose
handler has identity return case alone and the environment is empty.
\textsc{RetFib} returns value from a fiber to the previous one, evaluating the
body of the return case.

The rule \textsc{Handle} installs a handler by pushing a fiber with no frames
and the given handler. The rule \textsc{ExHn} handles an exception, if the
current handler has a matching exception case, unwinding the current fiber. The
rule \textsc{ExFwdC} forwards the exception to C. Here, there is exactly one
fiber on the current stack, and the handler does not have a matching exception
case, which we know is the case (see \textsc{Callback} rule). The rule
\textsc{ExFwdFib} forwards the exception to the next fiber if the current
handler does not handle it.

The rule \textsc{EffHn} captures the handling of effects when the current
handler has a matching effect case. We evaluate the body of the matching case,
and bind the continuation parameter |r| to the captured continuation. Observe
that the captured continuation |k''| includes the current handler. Intuitively,
the handler wraps around captured continuation. This gives Multicore OCaml
effect handlers deep handler semantics~\cite{}. \textsc{EffFwd} forwards the
effect to the outer fiber, and extends the captured continuation |k''| in the
process. Recall that we do not capture C frames as part of a continuation. To
this end, \textsc{EffUnHn} models unhandled effect. If the effect bubbles up to
the top fiber, which we know does not have an effect case (see
\textsc{Callback} rule), we raise \textsc{unhandled} exception at the point
where the corresponding effect was performed. This is achieved by appending the
captured continuation to the front of the current continuation.

Observe that |continue| and |discontinue| are not part of the expressions. They
are encoded as follows:
\begin{smathpar}
  \begin{array}{rcl}
    \textsf{continue} ~k ~e & = & (k ~(\olam{x}{x})) ~e \\
    \textsf{discontinue} ~k ~l ~e & = & (k ~(\olam{x}{\throw{l}{x}})) ~e
  \end{array}
\end{smathpar}
Intuitively, resuming a continuation in both the cases involves evaluating the
appropriate abstraction on top of the continuation. We perform the
administrative reductions \textsc{Resume1} and \textsc{Resume2} to evaluate the
arguments to |continue| and |discontinue|. The rule \textsc{Resume} appends the
given continuation to the front of the current continuation, and evaluates the
body of the closure.

\kc{Include example reductions from hl-semantics.ml in Appendix}

@@

In this section, we present the implementation details of effect handlers in
Multicore OCaml. The implementation follows the operational semantics, but has
a few key representational differences. Multicore OCaml currently only supports
the x86-64 backend on Linux, but our design preclude other backends. The
description in the sequel assumes we are on an x86-64 architecture.

Unlike the operational semantics, handlers with just exception patterns
(exception handlers) are implemented differently from effect handlers. As
mentioned in \S~\ref{sec:exn_handlers}, exceptions are pervasive in OCaml and
are so cheap that they are used for local control flow. Hence, we retain the
linked exception handler frame implementation of stock OCaml exception handlers
in Multicore OCaml for performance backwards compatibility. This differs from
other research languages with effect handlers~\cite{Links,Frank,Eff}, which
implement exceptions using effects (by ignoring the continuation argument in
the handler).

\begin{figure}
  \includegraphics[width=0.7\linewidth]{figures/fiber}
  \caption{Layout of Multicore OCaml fiber}
  \label{fig:fiber}
\end{figure}

\subsection{Heap-allocated fibers}

In the operational semantics, the continuations may be resumed more than once.
Captured continuations are copies of the original fibers and resuming the
continuation copies the fibers and leaves the continuation as it is. In our use
case, continuations will be resumed at-most once, and copying fibers is
unnecessary and inefficient. Instead, Multicore OCaml allocates fibers on the
heap. Figure~\ref{fig:fiber} shows the layout of a suspended fiber in Multicore
OCaml. A fiber is said to be suspended if it is not the current (top-most)
fiber during OCaml execution.

At the bottom of the stack, we have the |handler_info|, which contains the
pointer to parent fiber, and the closures for the value, exception and effect
cases. The closures are created by the translation described in
\S\ref{sec:static_semantics}; Multicore OCaml supports exception patterns in
addition to effect patterns in the same handler. This is followed by a block to
support DWARF unwinding, whose details we will return to later. Then, there is
a top-level handler that invokes the exception handler closure |clos_hexn|
after switching to the parent fiber. This is followed by the |pc| of the code
that invokes value handler after switching to the parent stack. The stack is
laid out such that when the computation in the handler returns, control goes to
value handler.

We have the variable-sized area for the OCaml frames next. Since we want the
handlers to be as small as possible, this area is initially 16 words in length.
When the stack pointer |rsp| goes less than stack threshold (a field in the
|Caml_state| table), the stack is said to have overflown. On stack overflow, we
copy the whole fiber to a new area with double the size. In Multicore OCaml, we
introduce stack overflow checks in function prologue of OCaml functions. We
observed that most function calls are to leaf functions with small frame size.
In order to take advantage of this fact, we introduce a small red zone at the
top of the stack. The compiler elides the stack overflow check for leaf
functions whose frame size is less than the size of the red zone.

Finally, we have the saved exception pointer, which points to the top-most
exception frame, and the saved stack pointer, which points to the top of the
stack. Switching between fiber only involves saving the exception and the stack
pointer of the current stack and loading the same on the target stack. We
allocate the space for the fiber using |malloc| and |free| it when no longer
required. We do not use a \emph{stack cache} of recently freed stacks for
speeding up fiber allocation. In our experiments, we observed that directly
using |mimalloc| outperforms having the stack cache.

Since C functions do not have stack overflow checks, we have to execute the
external calls in the system stack. Calling a C function from OCaml involves
saving the exception pointer and the stack pointer in the |fiber_info| of the
current fiber, and the allocation pointer in the |Caml_state|, updating |rsp|
to the top of system stack (maintained in |Caml_state|), and calling the C
function. The actions are reversed when returning from the external call. One
subtlety is that for C functions that take arguments on the stack, the
arguments must be copied to the C stack from the OCaml stack where they are
laid out. For vast majority of external calls, we perform only a few additional
instructions stock OCaml in order to to set up the right stack.

\begin{figure}
  \includegraphics[width=0.69\linewidth]{figures/multicore_stack}
	\caption{Layout of Multicore OCaml stack on \texttt{meander} example from
	\S\ref{sec:stack}.}
  \label{fig:mcstack}
\end{figure}

When we first enter OCaml from C, a new fiber is allocated for the main OCaml
stack. Since callbacks may be frequent in OCaml programs that use finalisers,
we run the callbacks on the same fiber as the current one. For example, the
layout of the Multicore OCaml stack at |caml_raise_exn| in the |meander|
example from \S\ref{sec:stack} is shown in Figure~\ref{fig:mcstack}. The
functions |caml_call_c| and |caml_call_ocaml| switch the stacks, and hence are
show in both the system stack and the fiber. The details of the exception
frames have been elided in the figure. Thanks to the fiber representation,
external calls and callbacks remain fast.
